A Bayesian statistical model is able to predict target-by-target selection behaviour in a human foraging task

### 2022-10-14 manuscriptRevisions2.tex
first pass at review comments

To assess how well our model can account for behaviour at the level of individual trials, we start by stepping through each trial in the data from Clarke and colleagues \cite{clarke2022}. For each target selection\footnote{Note: for the reasons discussed above, we ignore the initial target selection for now, as the model has very little to say about it. We return to this issue in section 4.}, in each trial, we can estimate how likely each of the remaining items are to be selected using the parameters from our posterior model. As can be seen in Figure \ref{fig:cal}, the model is putting somewhere between 25\% and 75\% of the weight on the target that is selected next, easily outperforming a chance $1/n$ baseline. We can also see  that the model is well calibrated in that the probabilities assigned to the most likely target manage to capture how often that target is actually selected. However, we can see a decrease in accuracy when it comes to how often ``runner-up`` candidate items are selected as our model appears to systematically undervalue these (presumably by putting too much of the probabilistic weight on low chance items).

@@

For trials in which the model does a worse job of predicting (accuracy <50$\%$) we can see that the human behaviour appears less organised in general and it seems unlikely that adding some form of path-length optimisation to our model would improve accuracy with these participants. We can see this more clearly in Figure \ref{fig:qjep_path_lengths}: while our model does a good job of capturing the average total path length, there is considerable variation with some participants generating shorter paths and some longer. We can see  that this appears to be systematically related to how well our model can predict behaviour.

@@

In our original model of visual foraging \citep{clarke2022foraging}, we were able to robustly measure aggregate parameters that underlie behaviour, at both an overall mean level across an experiment, and at the level of individual participants. In the current manuscript, we asked a more difficult question: could this model make target-by-target predictions, guessing which target would be picked next within a trial? Despite this being a relatively difficult task, we found that for some trials our model was over 80\% accurate, demonstrating that at least in some cases, participants' behaviour in this task is remarkably predictable.

@@

Our exploratory analyses seem to hint at the idea that the people who are difficult for the model to predict are those who do not behave in a manner that optimises path length. It could be that these participants are simply unpredictable, and do not behave in a consistent manner on a target-by-target basis, making it difficult for a model to predict their behaviour at this level of granularity. However, it is possible that there are other factors that we have not taken into account that could further help explain these discrepancies. For example, it is possible that relatively unpredictable participants use multiple rules and switch between them in ways that we are not currently captured in the model. Some participants may also lose track of where they are more frequently (perhaps due to attentional lapses) and thus may show foraging patterns that are predictable apart from unpredictable discontinuities, which again may not be easy to model. Another possibility is that participants may make exploratory, information-seeking visits to different parts of the array due to uncertainty about the targets present: this may be particularly important for conjunction trials, where peripheral visual information about shape is likely to be poorer. Finally, the displays had artificial discontinuities (i.e. screen edges) which may lead to different behaviour in different participants: some may carry out the task as if they are reading, reaching the right hand edge and then swinging back to the left, whereas others may 'bounce' off an edge they have just reached. Our model may well be able to predict the latter, but make less good predictions for the former behaviour.

>>>>

To assess how well our model can account for behaviour at the level of individual trials, we start by stepping through each trial in the data from Clarke and colleagues \cite{clarke2022}. For each target selection\footnote{Note: for the reasons discussed above, we ignore the initial target selection for now, as the model has very little to say about it. We return to this issue in section 4.}, in each trial, we can estimate how likely each of the remaining items are to be selected using the parameters from our posterior model. As can be seen in Figure \ref{fig:cal}, the model is putting somewhere between 25\% and 75\% of the weight on the target that is selected next, easily outperforming a chance $1/n$ baseline. There are also interesting differences between the two conditions: for example, in the conjunction condition, there is a clear 'jump' around target 20, which probably reflects the tendency for participants to forage in runs, selecting all exemplars of one target and then all exemplars of the other: there is a decrease in accuracy as people switch and then an increase again. We can also see that the model is well calibrated in that the probabilities assigned to the most likely target manage to capture how often that target is actually selected. However, we can see a decrease in accuracy when it comes to how often ``runner-up`` candidate items are selected as our model appears to systematically undervalue these (presumably by putting too much of the probabilistic weight on low chance items).

@@

For trials in which the model does a worse job of predicting (accuracy <50$\%$) we can see that the human behaviour appears less organised in terms of proximity in general and it seems unlikely that adding some form of path-length optimisation to our model would improve accuracy with these participants. We can see this more clearly in Figure \ref{fig:qjep_path_lengths}: while our model does a good job of capturing the average total path length, there is considerable variation with some participants generating shorter paths and some longer. We can see  that this appears to be systematically related to how well our model can predict behaviour. Interestingly, these poorly predicted example trials appear to have very strong pS biases, collecting all exemplars of one target and then the other: thus, the participants are carrying out a strategy that the model is not capturing well. This fits with conjunction trials being harder for the model to predict, as these trials generally have fewer switches.

@@

In our original model of visual foraging \citep{clarke2022foraging}, we were able to robustly measure aggregate parameters that underlie behaviour, at both an overall mean level across an experiment, and at the level of individual participants. In the current manuscript, we asked a more difficult question: could this model make target-by-target predictions, guessing which target would be picked next within a trial? We found that for some trials our model was over 80\% accurate, demonstrating that at least in some cases, our model is able to make good predictions of target behaviour.

@@

Our exploratory analyses seem to hint at the idea that the people who are difficult for the model to predict are those who do not behave in a manner that optimises path length. It could be that these participants are simply unpredictable, and do not behave in a consistent manner on a target-by-target basis, making it difficult for a model to predict their behaviour at this level of granularity. However, it is possible that there are other factors that we have not taken into account that could further help explain these discrepancies. Participants who 'stick' to one type of target are likely to have longer path lengths, and it seems that the model does do a poorer job of predicting conjunction trials, which generally have stronger pS biases. Thus, one area for future improvement of the model is to try to capture this behaviour better: it may be that a 'floor' parameter (on a participant-by-participant basis) in the spatial fall off, allocating more weight to more distant targets, may help account for currently difficult to predict trial behaviour.

There are also many other factors that the model does not account for. For example, it is possible that relatively unpredictable participants use multiple rules and switch between them in ways that we are not currently captured in the model. Some participants may also lose track of where they are more frequently (perhaps due to attentional lapses) and thus may show foraging patterns that are predictable apart from unpredictable discontinuities, which again may not be easy to model. Another possibility is that participants may make exploratory, information-seeking visits to different parts of the array due to uncertainty about the targets present: this may be particularly important for conjunction trials, where peripheral visual information about shape is likely to be poorer. Finally, the displays had artificial discontinuities (i.e. screen edges) which may lead to different behaviour in different participants: some may carry out the task as if they are reading, reaching the right hand edge and then swinging back to the left, whereas others may 'bounce' off an edge they have just reached. Our model may well be able to predict the latter, but make less good predictions for the former behaviour.
